---
title: "Disabling auto formatting feature for afuns in analyze setup"
author: "Gabriel Becker and Ilse Augustyns"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Disabling auto formatting feature for afuns in analyze setup}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}

.sourcecode {
  background-color:lightblue;
}
.reveal .r code {
    white-space: pre;
}
```
## Note
This vignette is based upon a version of `rtables` and `tern` that is not yet released. See here installation instructions for both packages.

```{r, echo=FALSE, eval=FALSE}
devtools::install_github(repo = "iaugusty/rtables", ref = "no_auto_fmt_alternative", force = TRUE, upgrade = "never")
devtools::install_github(repo = "iaugusty/tern", ref = "no_auto_fmt_alternative", force = TRUE, upgrade = "never")

```

## Setup

In the below we perform some setup to be utilized in the remainder of the vignette.

```{r, message=FALSE}
library(rtables)
library(tern)
library(dplyr)

ex_advs2a <- ex_advs %>%
  filter(PARAMCD %in% c("DIABP", "PULSE")) %>%
  select(ARMCD, PARAMCD, AVAL,SEX)

format_set1 <- c(n = "xx", dumb = "xx.x", mean = "xx.xxx", range = "xx.xx - xx.xx", count_fraction = "xx (xx.xx%)")
format_set2 <- c(n = "xx", dumb = "xx.x", mean = "xx.xxx", range = "xx.xx - xx.xx", count_fraction = junco::jjcsformat_count_fraction)



```



## Introduction: Need for varying format specifications per facet

Here we describe 2 basic scenarios that we often encounter, where we would like to easily control the behavior of the format specifications per different facet.

The first situation where we’d like to update the formats according to facet we are in is demonstrated with the following example. 
We have a table with different parameters (here `DIABP` and `PULSE`), in which we like the formatting specifications being different for these 2 parameters.

In the below layout, we only can specify a fixed formatting, ie the same `.formats` is being applied to all facet splits within the table.

Our goal is to avoid the need for rewriting afuns that can deal with adjusting formats according to which split you are in.


```{r}
lyt1 <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze(
    vars = "AVAL",
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range"),
      .formats = c(n = "xx", mean = "xx.xx", range = "xx.xx - xx.xx")
    )
    
  )

tbl1 <- build_table(lyt1, ex_advs2a, ex_adsl)
tbl1
```
Another common situation where we’d like to update the formats according to facet we are in is described with the following example. 

We have a table with different variables being analyzed (here `AGE`, `BMRKR1`, `STRATA1`, `EOSDY`), in which we like the formatting specifications being different for these 4 variables.

In the below layout, at least with one call to `analyze`, we only can specify a fixed formatting, ie the same `.formats` is being applied to all facet splits (here variables) within the table.


```{r, message=FALSE}
lyt2 <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  analyze(
    vars = c("AGE", "BMRKR1", "STRATA1", "EOSDY"),
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range", "count_fraction"),
      .formats = format_set1
    )
  )

tbl2 <- build_table(lyt2, ex_adsl)
tbl2
```

For this table, there is obviously the option to construct the table with 4 different calls to `analyze`, and in each call the desired `.format` argument is specified. 

Our goal is to avoid the need for calling `analyze` for each variable separately or rewriting afuns that can deal with adjusting formats according to which split you are in.

Note that above 2 layout definitions make use of `tern` `afun` `a_summary`. A note on this later.

## new feature to analyze function to turn off auto formatting

As a solution to the above problem, a new `rtables` feature, called `no_auto_fmt`, has been introduced.

This `no_auto_fmt` approach can be applied to afuns that take the arguments `.formats` and `.stats` to specify statistics and formats to control formatting of requested statistics.

All `tern` afuns are written in this way. In addition to that, the argument `.formats`, when incomplete or `NULL`, is being updated inside the afun to always end up with formatting specification passed onto an `in_rows` call. This is what we refer to as `auto_fmt` or auto formatting handling.

The goal of this new feature is to be able to turn off this automated formatting handling when calling these afuns.

The idea is that, when `.formats` = `no_auto_fmt`, the automatic updating is turned off, and the `format` argument from `analyze` can be used to control the formats of the requested statistics.


Let's start with the following simple table, a simplified version of the second scenario. 
Note that in below layout definition the `.format` argument is not specified, implying `tern` auto formatting feature is still on.
In this case, the `analyze` `format` argument is ignored.
```{r, message=FALSE}

lyt2x <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  analyze(
    vars = "AGE",
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range")
    ),
    format = format_set2
  )

tbl2x <- build_table(lyt2x, ex_adsl)
tbl2x

```

For purpose of comparison with a later table, the same table is regenerated with a user defined `.formats` argument. 
Again, the `format` argument to `analyze` is being ignored.

```{r, message=FALSE}

lyt2a <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  analyze(
    vars = "AGE",
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range"),
      .formats = format_set1
    ),
    format = format_set2
  )

tbl2a <- build_table(lyt2a, ex_adsl)
tbl2a

```

### Simply turning off auto formatting feature, showing raw values

In the below example, as a first step the formatting from within `afun` is turned off, which is done by setting `.formats` = `no_auto_fmt`.
As `format` = `NULL`, no formatting will be applied, and raw values are being presented.
This is not yet how users would apply the new features, it is just to demonstrate the steps in the process.

```{r, message=FALSE}
lyt2b <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  analyze(
    vars = "AGE",
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range"),
      .formats = no_auto_fmt
    ),
    format = NULL
  )

tbl2b <- build_table(lyt2b, ex_adsl)
tbl2b
```

### Turning off auto formatting feature, with format argument to show formatted values, but not yet per facet

In the below example, as a second step in the process, note the `format` argument to `analyze` is now a regular format specification, which is used in the table creation.
The current example is not yet how users would apply the new features either, as the outcome is still the same as the original situation with `.formats` = format_set1 (see earlier example).

```{r, message=FALSE}
lyt2c <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  analyze(
    vars = "AGE",
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range"),
      .formats = no_auto_fmt
    ),
    format = format_set1
  )

tbl2c <- build_table(lyt2c, ex_adsl)
tbl2c


identical(get_formatted_cells(tbl2a), get_formatted_cells(tbl2c))
```


### Turning off auto formatting feature, with format argument to show formatted values different formats accross facets
#### tern function a_summary

As a final step in the process towards our goal of different formats across different facets, look at the following layout specification.

First take a look at the following `format_list` specification. 
`format_list` is a list specification setup for specific variables (and an extra `default` to cover unspecified variables).

```{r}
format_list <- list(
  "AGE" = c(n = "xx", mean = "xx.x", range = "xx - xx", dumb = "xx.x", "count_fraction" = "xx (xx.xx%)"),
  "BMRKR1" = c(n = "xx", mean = "xx.xxxx", range = "xx.xx - xx.xx", dumb = "xx.x", "count_fraction" = "xx (xx.xx%)"),
  "default" = c(n = "xx", mean = "xx.xx", range = "xx.xx - xx.xx", "count_fraction" = "xx (xx.xx%)")
)

format_list
```

Now take a look at the beow layout, which uses `analyze`, `afun` = `a_summary`, `.formats` = `no_auto_fmt` and format = `format_list`. 

This example fulfills our need to be able to have easy approach for providing variable specific formatting specifications to analyze multiple variables together, see initial version `lyt2` as the starting point.

```{r}
lyt2d <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  analyze(
    vars = c("AGE", "SEX", "BMRKR1", "STRATA1", "EOSDY"),
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range", "count_fraction"),
      .formats = no_auto_fmt),
    format = format_list
  )

tbl2d <- build_table(lyt2d,  ex_adsl)
tbl2d
```
As you can see in the above output, we succeeded in our goal to have different formats for each of the different variables presented in the table. 

Now we continue with the first layout goal: facets in the table coming from multiple parameters rather than multiple variables (see `lyt1` as the starting point).


```{r, message=FALSE}

ex_advs2 <- ex_advs2a %>%
  mutate(fmt_col = case_when(
    PARAMCD == "DIABP" ~ list(c(n = "xx", dumb = "xx.x", mean = "xx.xx", range = "xx.x - xx.x", lsmean = "xx.xx")),
    PARAMCD == "PULSE" ~ list(c(n = "xx", dumb = "xx.x", mean = "xx.xxx", range = "xx.xx - xx.xx", lsmean = "xx.xxx"))
  )) 



```

As part of the process to specify formats per `PARAMCD`, we have extended the input dataframe with a column `fmt_col`, in which the formatting specifications are provided per `PARAMCD`.
For `PARAMCD` = `DIABP`, the specifications are to present mean with 2 decimals, while for `PARAMCD` = `PULSE`, the specifications are to present mean with 3 decimals. 



```{r, message=FALSE}

    fmt_spec <- unique(ex_advs2 %>% 
      select(PARAMCD, fmt_col))
    fmt_spec
    fmt_spec[["fmt_col"]]
  
```
Similar as before, turning off the tern auto formatting is done by `.formats` = `no_auto_fmt` and the parameter dependent formatting specifications are handled by `analyze` argument `format` = "fmt_col". 

```{r, message=FALSE}

lyt1a <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze(
    vars = "AVAL",
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range"),
      .formats = no_auto_fmt
    ),
    format = "fmt_col"
  )

tbl1a <- build_table(lyt1a, ex_advs2, ex_adsl)
tbl1a
```

The same code, but without `.formats` = `no_auto_fmt` and providing `format` = "fmt_col" to `analyze` call, still would result in the original `tern` behaviour, ie fixed formats for all `PARAMCD`.

```{r, message=FALSE}
lyt1b <- basic_table() %>%
  split_cols_by("ARMCD") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze(
    vars = "AVAL",
    afun = a_summary,
    extra_args = list(
      .stats = c("n", "mean", "range")
    ),
    format = "fmt_col"
  )

tbl1b <- build_table(lyt1b, ex_advs2, ex_adsl)
tbl1b
```

#### other tern functions eg a_ancova
In the above examples we have discussed so far, we were using the `tern` `afun`  `a_summary`. 
Note that not all `tern` afuns can be used in `analyze` call, see a short explanation on this later in this vignette.

As a general rule, instead, the **analyze functions** need to be used.
The **analyze function** associated with `a_summary` is `analyze_vars`.
Another `tern` **analyze function** is `summarize_ancova`, which is based upon the `afun` `tern:::a_ancova`.
See list of `tern` **analyze functions** [here](https://insightsengineering.github.io/tern/latest-tag/reference/analyze_functions.html).

Here we demonstrate, that the `no_auto_fmt` approach can be utilized for these analyze functions.
Again, the `tern` auto formatting is turned off by `.formats` argument, and the formatting is taken from the `format` argument to `analyze`.

```{r, message=FALSE}
lyt5a <- basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze_vars(
    vars = "AVAL",
    table_names = "table 1",
    var_labels = "Analyze_vars",
    conf_level = 0.95,
    .stats = c("n", "mean"),
    .formats = no_auto_fmt,
    format = "fmt_col" 
  ) %>% 
  summarize_ancova(
    vars = "AVAL",
    table_names = "table 2",
    variables = list(arm = "ARMCD", covariates = "SEX"),
    var_labels = "summarize_ancova",
    conf_level = 0.95,
    .stats = c("n", "lsmean"),
    .formats = no_auto_fmt,
    format = "fmt_col" 
  )

tbl5a <- build_table(lyt5a, ex_advs2, ex_adsl)
tbl5a
```

Similar as before, when the `tern` auto formatting is not turned off by `.formats` argument, the `format` argument to `analyze` is being ignored.

```{r, message=FALSE}

lyt5b <- basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze_vars(
    vars = "AVAL",
    table_names = "table 1",
    var_labels = "Analyze_vars",
    conf_level = 0.95,
    .stats = c("n", "mean"),
    format = "fmt_col" 
  ) %>% 
  summarize_ancova(
    vars = "AVAL",
    table_names = "table 2",
    variables = list(arm = "ARMCD", covariates = "SEX"),
    var_labels = "summarize_ancova",
    conf_level = 0.95,
    .stats = c("n", "lsmean"),
    format = "fmt_col" 
  )

tbl5b <- build_table(lyt5b, ex_advs2, ex_adsl)
tbl5b
```




#### a note on using tern afuns directly in analyze call
As noted before in the above examples we have switched to using `tern` analyze functions (`summarize_vars` and `summarize_ancova`), not afuns.


Here we demonstrate that also the `tern` afuns can be utilized directly in `analyze` calls, when utilizing the `no_auto_fmt` feature.

We start again with the basic `a_summary` `afun` in one call and the `tern:::a_ancova` function in another `analyze` call.
Note the usage of `.formats` = `no_auto_fmt` in both calls.

```{r, message=FALSE}
lyt6a <- basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze(
    vars = "AVAL",
    afun = a_summary,
    table_names = "SUMMARY",
    var_labels = "My label- summary",
    extra_args = list(
      .stats = c("n", "mean", "range"),
      .formats = no_auto_fmt
    ),
    format = "fmt_col"
  ) %>%
  analyze(
    vars = "AVAL",
    afun = tern:::a_ancova,
    table_names = "ANCOVA",
    var_labels = "My label- ANCOVA",
    extra_args = list(
      variables = list(arm = "ARMCD", covariates = "SEX"),
      conf_level = 0.95,
      .stats = c("n", "lsmean"),
      .formats = no_auto_fmt
  ),
  format = "fmt_col")

tbl6a <- build_table(lyt6a, ex_advs2, ex_adsl)
tbl6a
```

Note that it is not possible to not disabling auto formatting and using an `analyze` call with a `tern` `afun` directly, unless the `afun` takes `x` as the first argument (like `a_summary` does).

The below code (the part to `afun` = `tern:::a_ancova`) does not run and yields the following error : 

`Error: Error applying analysis function (var - AVAL): argument ".var" is missing, with no default`

```{r, message=FALSE, eval = FALSE}
lyt6b <- basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze(
    vars = "AVAL",
    afun = a_summary,
    table_names = "SUMMARY",
    var_labels = "My label- summary",
    extra_args = list(
      .stats = c("n", "mean", "range")
    )
  ) %>%
  analyze(
    vars = "AVAL",
    afun = tern:::a_ancova,
    table_names = "ANCOVA",
    var_labels = "My label- ANCOVA",
    extra_args = list(
      variables = list(arm = "ARMCD", covariates = "SEX"),
      conf_level = 0.95,
      .stats = c("n", "lsmean")
  ))

tbl6b <- build_table(lyt6b, ex_advs2, ex_adsl)
tbl6b
```


A solution to that is to apply the new `update_afun_no_auto` function to the underlying `tern` `afun`, and to utilize that updated function in the `analyze` call.


```{r, message=FALSE, eval=TRUE}
myafun <- update_afun_no_auto(afun = tern:::a_ancova)


lyt6c <- basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  split_rows_by("PARAMCD", split_fun = drop_split_levels) %>%
  analyze(
    vars = "AVAL",
    afun = a_summary,
    table_names = "SUMMARY",
    var_labels = "My label- summary",
    extra_args = list(
      .stats = c("n", "mean", "range")
    )
  ) %>%
  analyze(
    vars = "AVAL",
    afun = myafun,
    table_names = "ANCOVA",
    var_labels = "My label- ANCOVA",
    extra_args = list(
      variables = list(arm = "ARMCD", covariates = "SEX"),
      conf_level = 0.95,
      .stats = c("n", "lsmean")
    ))

tbl6c <- build_table(lyt6c, ex_advs2, ex_adsl)
tbl6c

```

#### background for tern afuns 

As indicated earlier, it is not possible to directly utilize `tern` afuns that take `df` as the first argument inside an `analyze` call.

The reason for this is how the afuns are being defined. The specialized arguments (see `?additional_fun_params`) are not named in formals, and go anonymously into `...` instead. Due to this reason, the splitting machinery cannnot be directly applied to these afuns.

Inside the corresponding **analyze function** (eg `summarize_ancova` corresponds to afun `tern:::a_ancova`), the underlying `afun` is updated in order to retrieve the facet values for these specialized arguments.

A similar technique has been used in the current solution of `no_auto_fmt`.

This technique has an impact on debugging methods. See next section.









